---
title: "Quantile dotplots"
output: 
    github_document:
        toc: true
---

## Introduction

This document describes *quantile dotplots*, the basic motivation behind them, and how to generate them.

Please cite:

Matthew Kay, Tara Kola, Jessica Hullman, Sean Munson. _When (ish) is My Bus? 
User-centered Visualizations of Uncertainty in Everyday, Mobile Predictive Systems_. 
CHI 2016. DOI: [10.1145/2858036.2858558](http://dx.doi.org/10.1145/2858036.2858558).

## Setup

### Required libraries

If you are missing any of the packages below, use `install.packages("packagename")` to install them.
The `import::` syntax requires the `import` package to be installed, and provides a simple way to 
import specific functions from a package without polluting your entire namespace (unlike `library()`)

```{r setup, results="hide", message=FALSE}
library(ggplot2)
import::from(magrittr, `%>%`, `%<>%`, `%$%`)
import::from(dplyr, 
    transmute, group_by, mutate, filter, select, data_frame,
    left_join, summarise, one_of, arrange, do, ungroup)
```

### Ggplot theme

```{r setup_ggplot}
theme_set(theme_light() + theme(
    panel.grid.major=element_blank(), 
    panel.grid.minor=element_blank(),
    axis.line=element_line(color="black"),
    text=element_text(size=14),
    axis.text=element_text(size=rel(15/16)),
    axis.ticks.length=unit(8, "points"),
    line=element_line(size=.75)
))
```

## Representing a continuous probability distribution as discrete outcomes

We might like to represent a continuous probability distribution (say, a prediction for when a bus is going to arrive) as discrete outcomes, since frequency-based (or "discrete outcome") presentations can be easier for people to interpret. How should we do that?

### A problematic solution: random draws

A first pass at representing a continous probability distribution as discrete outcomes might be to generate random draws from that distribution. However, especially for a small number of samples, this tends not to work well. The problem is that any given random sample of (say) 20 isn't always representative.

For example, a log-normal distribution might look like this:

```{r reference_dist, fig.width=7, fig.height=2.5}
mu = log(11.4)
sigma = 0.2
x = seq(from=.01, to=30, length=10001)
d = dlnorm(x, mu, sigma)
density_df = data_frame(x, d)
density_df %>%
    ggplot(aes(x = x, y = d)) +
    geom_line() +
    geom_vline(xintercept = 11, color="red") + 
    xlim(0, 30)
```

And we could generate random samples of a set size, say 20, to visualize this. But every time we
do that it looks different:

```{r random_draws, fig.width=7, fig.height=11, warning=FALSE}
runs = 5
samples = 20
binwidth = 1.25
data.frame(
    run = factor(1:runs),
    x = rlnorm(runs * samples, mu, sigma)
) %>%
    ggplot(aes(x=x)) +
        geom_dotplot(binwidth=binwidth) +
        facet_grid(run ~ .) +
        xlim(0, 30)
```

Sometimes the distribution shape is obscured, the location shifted, or tails over- or under-represented.

### A consistent solution: a quantile dotplot

One way to get a more consistent representation is to use the quantile function of the distribution instead of random draws. In essence, this encodes the cumulative distribution function
one-dimensionally, and then turns it into a Wilkinsonian dotplot. (Wilkinson originally described dotplots for displaying sample data in [Wilkinson, L, Dot Plots, _The American Statistician_, 1999](dx.doi.org/10.1080/00031305.1999.10474474); we re-use them here for displaying predictive quantiles, hence _quantile dotplot_).

First, we generate evenly-space quantiles in probability space (i.e. from 0 to 1) depending on the number of `samples` we want. If you are familiar with Q-Q plots, this is essentially the same approach used to generate representative quantiles for a Q-Q plot (a variant of this method is implemented in R in the `ppoints` function, but spelled out here for completeness).

```{r quantiles}
quantiles = data_frame(
    p_less_than_x = seq(from = 1/samples / 2, to = 1 - (1/samples / 2), length=samples),
    x = qlnorm(p_less_than_x, mu, sigma)
)
quantiles
```

The table above shows the probability of drawing a value less than x (i.e. `P(X < x)`) and the corresponding value of x to achieve that probability on the underlying distribution. We generate 20 (or whatever value of `samples` you like) values of `p_less_than_x` evenly-spaced in `(0,1)` (i.e. not including 0 or 1), and then find `x` using the inverse CDF (aka the quantile function) of the predictive distribution at each value of `p_less_than_x`. If we then take those values of `x` and plot them as a dotplot, we get a quantile dotplot:

Visualized with the CDF, that looks like this:

```{r quantile_dotplot_generation, fig.width=7, fig.height=3, fig.show="hold"}
x = seq(0.01, 30, length=1001)
p_less_than_x = plnorm(x, mu, sigma)
qf_df = data_frame(x, p_less_than_x)
cdf_plot = ggplot(quantiles, aes(x=x, y=p_less_than_x)) +
    geom_segment(aes(xend=x, yend=p_less_than_x), x = 0, color="gray75", linetype="dashed") +
    geom_segment(aes(xend=x, yend=p_less_than_x), y = -0.05, color="gray75", linetype="dashed",
        arrow = arrow(ends="first", length=unit(5, "pt"), type="closed")) +    
    geom_line(data = qf_df, color="red", size=1) +
    annotate("text", x = 30, y = 1, label = "cumulative distribution function", color="red", hjust=1, vjust=1.5) +
    xlim(c(-3.5, 30))
cdf_plot

quantile_dotplot = quantiles %>%
    ggplot(aes(x = x)) +
    geom_vline(aes(xintercept = x), color="gray75", linetype="dashed") +  
    geom_dotplot(binwidth = 1.25, color=NA) +
    xlim(-3.5,30)
quantile_dotplot
```

The evenly-spaced probabilities on the y axis are turned into representative quantiles from the distribution on the x axis. 

## Properties of the quantile dotplot

### Shape approximates density
The shape a quantile dotplot approximates the shape of the density:

```{r quantile_dotplot_with_density, fig.width=7, fig.height=2}
quantile_dotplot +
    geom_line(aes(y = d * 5.1), data = density_df, color="red")
```

### Finding probability intervals reduces to counting

Inferences about one-sided probability intervals on the distribution can
be made through counting. If our example distribution represents a probability
distribution over predicted time to arrive for my bus, and I am willing to miss
the bus 2/20 times, I can count up 2 dots ("busses") from the left to get the
time I should arrive at the bus stop. This works because the x-value for 
`P(X < x) = 2/20` will be between the second and third dot. More generally, the x-value for
`P(X < x) = k/20` will be between the `k`th and `k + 1`th dot). Here's how this works
on the CDF and the dotplot:

```{r counting_on_quantile_dotplot, fig.width=7, fig.height=3, fig.show="hold"}
cdf_plot +
    geom_segment(x = 0, xend = qlnorm(.1, mu, sigma), y = .1, yend = .1, color = "purple", size = 1) +  
    geom_segment(x = qlnorm(.1, mu, sigma), xend = qlnorm(.1, mu, sigma), y = -0.05, yend = .1, 
        color = "purple", size = 1, arrow = arrow(ends="first", length=unit(5, "pt"), type="closed")) +
    annotate("text", x = 0, y = 0.1, label = "2/20 = 0.1", color = "purple", hjust = 1.1)
quantile_dotplot +
    geom_dotplot(binwidth = 1.25, color=NA, fill="purple", data=filter(quantiles, p_less_than_x < .1)) +
    geom_vline(xintercept = qlnorm(.1, mu, sigma), color="purple", size=1) 
```    

Consequently, this purple line represents the point of time I should arrive at my bus if I want to catch it about 18 times out of 20 (or miss it 2 times out of 20).
